---
title: "Real data analysis"
format: html
editor: visual
---

```{r}
rm(list = ls())
# load libraries
library(rlang, lib.loc = "/home/qianzhao_umass_edu/R/x86_64-pc-linux-gnu-library/4.5/")
library(tidyverse)
library(gridExtra)
library(glmnet)
library(xgboost)
library(ggpubr)
```

This document contains all the code used to analyse the PAISA data using the tilted knockoff method. We analyze three secondary phenotype: EID, ATT, and PS. For each of these traits we separately analyze two measures: accuracy and speed. The observed values are turned into binary by setting the top 20% as 1 and the rest as 0.

```{r}
# all diagnoses and phenotypes
diagnoses <- c("BP1", "BP2", "MDD", "SCZ","Control")
phenos <- c("EID_A_Binary", "EID_S_Binary", "ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary")
```

```{r}
# Incidence of each diagnosis
incidence_rates <- list(
  BP1 = 219 / 2 / 100000,
  BP2 = 219 / 2 / 100000,
  MDD = 235 / 100000,
  SCZ = 45 / 100000,
  Control = 0.99501
)
```

# Pre-processing the genotype data

```{r}
data_dir <- "directory_to_raw_data"
```

Because the raw data is large (containing all the SNPs), we first obtain a subset of data containing the SNPs that have been found to be related to severe mental illnesses in previous studies.

### Extract the SNP names

First, I identify a list of candidate SNPs, these are the SNPs that have been found to be associated with at least one of the diagnosis.

```{r get_snp_names}
geno <- read_plink(file = paste0(data_dir, "9KSmallSNP11282023/GWASIndexSNPs"))
snp_names <- geno$bim
```

The list of SNP names is stored in the file `snp_names_12042023.csv`.

### Extract relevant SNPs from the raw data

Next, I extract measurements at these relevant SNPs for each individual. The following step uses Julia language.

```{r}
# Using julia code to get data 
using SnpArrays
using DelimitedFiles
using CSV
using DataFrames

# read snp names
df= CSV.read("/home/users/qzhao1/paisa/collider/real/snp_names_12042023.csv", DataFrame)

# location of the genotype data  
paisa_dir = "/oak/stanford/groups/candes/PAISA_new/" # original data (before QC)
des_dir = "/oak/stanford/groups/candes/PAISA/" # destination

chr = collect(1:22)

for i = 1:22
  # create row/col index
  xdata = SnpData(joinpath(paisa_dir, "chr" * string(chr[i])))
  
  # positions
  pos = filter(:chr => name -> name == chr[i], df).pos
  
  # create row/col indices to keep 
  snp_idx = indexin(pos, xdata.snp_info[!, "position"]) |> Vector{Int}
  
  # destination file names
  src = joinpath(paisa_dir, "chr" * string(chr[i]))
  des = joinpath(des_dir, "chr" * string(chr[i]) *"small") # plink file name 
  
  # filter
  SnpArrays.filter(xdata,  1:9007, snp_idx, des = des)
end 

# to run a for loop use for ...  end  (https://docs.julialang.org/en/v1/manual/control-flow/#man-loops)
```

A small data containing the SNPs in each chromosome is created after this step (they are stored in the folder `9KSmallSNPRaw12072023`).

### Concatenate chromosomes to a single data frame

Since raw data was separated into individual chromosomes, we now concatenate data into a single data frame. We use R for this step

```{r concatenate_chromosomes}
geno_raw <- NULL
subject_id_raw <- NULL # store the subject ID in chr 1 and then compare with all the other chromosomes to make sure that they all match

for(chr in 1:22){
  geno_raw_new <- read_plink(file = paste0(data_dir, "9KSmallSNPRaw12072023/chr", chr, "small"))
  new_bim <- geno_raw_new$bim
  new_bim$pos <- trimws(new_bim$pos)
  
  new_X <- geno_raw_new$X
  # paste together the genotype name
  new_snp <-  paste0("X", apply(new_bim[, c("chr", "pos", "alt", "ref")], 1, paste, collapse = "_"))
  
  rownames(new_X) <- new_snp
  if(chr == 1){
    subject_id_raw <- geno_raw_new$fam$id
    colnames(new_X) <- subject_id_raw
  }else{ # check if the individual ids match
    if(!identical(geno_raw_new$fam$id,subject_id_raw)) cat("Subject ID are different in chr", chr, "!")
  }
  
  geno_raw <- rbind(geno_raw, new_X)
}

geno_raw <- as.data.frame(geno_raw)
```

In the `geno_raw` data, each column contains the SNPs of an individual. The column names are individual IDs in the raw data. I map the individual IDs to their Paisa ID in the next step.

### Map individual IDs to their Paisa ID

```{r get_id_mapping}
PaisaIDMapping <- read_delim(paste0(data_dir, "/9KSmallSNP11282023/PaisaIDMapping.txt"),  delim = "\t", escape_double = FALSE, trim_ws = TRUE)
```

Obtain the PAISA ID fore each individual.

```{r map_paisa_id}
mapped_bid <- tibble(
  bid = colnames(geno_raw)
) %>% 
  left_join(PaisaIDMapping, by =  c("bid" = "BID")) 
```

Replace the ID in raw data with their Paisa ID.

```{r replace_id}
colnames(geno_raw) <- mapped_bid$PaisaID
geno_raw <- geno_raw[,which(!is.na(colnames(geno_raw)))]
geno_raw <- t(geno_raw)
geno_raw <- as.data.frame(geno_raw)
geno_raw$PaisaID <- rownames(geno_raw)
```

### Remove related individuals

We remove related individuals in the data, the remaining individuals are maximally related as first degree cousin in the data.

```{r remove_id}
id_toremove <- scan(paste0(data_dir, "9KPhenoData/ID_ToRemove/HaveCNB.Out_toRemove.txt"))

geno_raw <- geno_raw %>% 
  filter(!PaisaID %in% id_toremove)
```

### Store raw data

This raw data is stored at `Processed/processed_data/geno_raw.txt`

```{r store_geno_raw}
write.table(geno_raw, paste0(data_dir,  "Processed/processed_data/geno_raw.txt"),row.names = F, col.names = T)
```

### Remove highly correlated SNPs

To increase power of the knockoff approach, we remove highly correlated SNPs from the data.

#### Load raw genotype data

We start from the raw genotype data from previous step.

```{r load_geno_raw}
geno_raw <- read.table(paste0(data_dir, "Processed/processed_data/geno_raw.txt"), header = T)
```

#### Find pairs of SNPs with higher correlation

We set the threshold of correlation among pairs of SNPs to be $\rho = 0.8$. Suppose multiple SNPs are correlated with $\rho > 0.8$, we pick one representative among the group. For example, if SNP1 and SNP2, SNP2 and SNP3 have correlation above the threshold, we pick one representative among the three SNPs.

```{r}
thresh <- 0.8
```

```{r pairwise_corr}
# compute pairwise correlations
geno_raw_no_id <- select(geno_raw, -PaisaID)
snp_names <- colnames(geno_raw_no_id)
snp_corr <- cor(geno_raw_no_id, geno_raw_no_id, use = "everything") %>%
  as_tibble()
snp_corr <- snp_corr %>%  mutate(snp1 = colnames(snp_corr)) %>% 
  pivot_longer(cols = 1:ncol(snp_corr), names_to = "snp2", values_to = "corr") 

snp_corr_large <- snp_corr%>% 
  filter(snp1 != snp2)  %>% # remove rows that are the same 
  filter(corr > thresh) 
```

There are `r nrow(snp_corr_large)` above the threshold.

#### Merge multiple pairs into groups

```{r merge_snp_to_groups}
corr <- snp_corr_large
candidates <- sort(unique(c(snp_corr_large$snp1, snp_corr_large$snp2))) # indices of the SNPs that have strong correlations with each other 

group_list <- list()
ngroups <- 1

j <- 1 # index within a group
curr_snp <- candidates[1]
group_list[[ngroups]] <- curr_snp

flag <- F
while(1){
  # which candidates are related to the current snp
  indices <- which(corr$snp1 == curr_snp | corr$snp2 == curr_snp)
  if(length(indices) > 0){
    related_snp <- setdiff(c(corr$snp1[indices], corr$snp2[indices]), i)
    corr <- corr[-indices, ]
        
    # add this snp to the current group  
    group_list[[ngroups]] <- c(group_list[[ngroups]], setdiff(related_snp, group_list[[ngroups]]))
  }
  
  j <- j + 1
  if(j > length(group_list[[ngroups]])){ # are any new SNPs added to the group? 
    ngroups <- ngroups +1
    j <- 1
    curr_snp <- setdiff(candidates, unlist(group_list))[1]
    if(is.na(curr_snp)){
      flag = T
    }else{
      group_list[[ngroups]] <- curr_snp
    }
  }else{
    curr_snp <- group_list[[ngroups]][j]
    cat(curr_snp, ",")
  }
  if(flag == TRUE) break;
}
```

#### Pick the first SNP in each group as the representative

```{r snp_to_remove}
snp_to_remove <- lapply(group_list, function(t) return(t[-1])) %>% unlist() # remove SNPs that are not representatives 

snp_to_keep <- setdiff(snp_names, snp_to_remove)
```

```{r store_snp_for_analysis}
write.table(snp_to_keep, paste0(paste0(data_dir, "Processed/processed_data/snp_to_keep_0318.txt")), row.names = F, col.names = F)
```

Prune the genotype data to include only SNPs we keep for analysis.

```{r store_geno_pruned}
geno_pruned <- select(geno_raw, all_of(snp_to_keep), PaisaID)

write.table(geno_pruned, paste0(data_dir, "Processed/processed_data/geno_pruned_0318.txt"), quote = F, row.names = F, col.names = T)
```

### List of SNPs affecting each diagnosis

We compute SNPs that has been found in previous research to be associated to each diagnosis. For SNPs in a group, I replace them with the representative SNP.

```{r load_effect_size}
EffectSizeIndexSNPs <- read_table(paste0(data_dir, "9KSmall01112024/EffectSizeIndexSNPs.txt"))
```

```{r snp_per_diagnosis}
snp_to_keep <- read.table(paste0(data_dir, "Processed/processed_data/snp_to_keep_0318.txt"))$V1

for(diagnosis in diagnoses){
  if(diagnosis %in% c("BP1", "BP2")){
    snp_diagnosis <- EffectSizeIndexSNPs %>% filter(GWAS_file == "BD") %>% pull(sigSNPS)
  }else{
    snp_diagnosis <- EffectSizeIndexSNPs %>% filter(GWAS_file == diagnosis) %>% pull(sigSNPS)
  }
  
  snp_diagnosis <- paste0("X", snp_diagnosis)
  
  signal_snp_rep <- NULL
  for(i in 1:length(group_list)){
    if(length(intersect(snp_diagnosis, group_list[[i]]))!=0){
      signal_snp_rep <- c(signal_snp_rep, group_list[[i]][1])
    }
  }
  
  snp_diagnosis <- intersect(snp_diagnosis, snp_to_keep) # removes non-representative SNPs in a group
  snp_diagnosis <- unique(c(snp_diagnosis, signal_snp_rep)) # add group representatives 
  
  # store the signal SNPs for each diagnosis 
  write.table(snp_diagnosis, paste0(data_dir, "Processed/processed_data/snp_diagnosis_", diagnosis, "_0318.txt"), row.names = F, col.names = F)
}
```

### Summary of data

We summarize the data after processing that we will use for the analysis

-   `geno_pruned` - measurement of each genetic variants and PaisaID of each individual

-   `cnb` - binary Computerized Neurocognitive Battery (CNB) value for each secondary phenotype

-   `dx` - diagnosis and PaisaID of each individual

-   `id_toremove` - PaisaIDs of individuals who are related to another participant in the data

-   `snp_diagnosis` - relevant SNP for each diagnosis

These data are combined to a single data per phenotype (we do this because different individuals have different missing data for secondary phenotypes).

```{r separate_phenotype_data}
for(pheno in phenos){
  data <- left_join(geno, select(cnb, c(PaisaID, {{pheno}})), by = "PaisaID") %>% 
    filter(!PaisaID %in% id_toremove) %>% 
    filter_at(vars({{pheno}}), all_vars(!is.na(.))) %>% 
    left_join(dx, by = "PaisaID") 
  
  write.table(data, paste0(data_dir, "Processed/data_pheno/all_data_", 
                                     pheno, "_08092026.txt"), row.names = F, col.names = T)
}
```

# Variable selection

We consider four approaches:

-   IPW with logistic regression + sandwich estimator to compute uncertainty

-   Standard Knockoffs without adjustment

-   Case/control specific knockoffs

-   Tilted knockoffs

    For IPW and tilted knockoffs that requires an estimate of $P(D = d | X, Y)$ we use two approaches to estimate the model:

<!-- -->

-   The logistic regression to estimate $P(D = d | X, Y) / P(D = "Control" | X, Y)$ for each diagnosis (except for control) and then re-adjust the weights to get the conditional probability of each diagnosis.

-   XGBoost to estimate the probability of each diagnoses.

## Set up

```{r}
n_knockoff <- 10 
kappa <- 3  
```

Target FDR level

```{r}
q <- 0.3
```

Parameters for XGBoost

```{r}
xgb_params <- list(
  max_depth = 7,
  eta = 0.1,
  nrounds = 400,
  early_stopping_rounds = 10,
  subsample = 0.5
)
```

Directories

```{r}
data_dir <- "directory_to_data"
knockoff_dir <- paste0(data_dir, "ProcessedData/Knockoffs/")
```

### Load data

Load the known SNPs associated with each specific diagnosis

```{r}
snp_diagnosis_list <- list()
for(diagnosis in diagnoses) {
  if(diagnosis == "Control") next
  snp_diagnosis_list[[diagnosis]] <- 
    read.table(paste0(data_dir, "snp_diagnosis/snp_diagnosis_",
                      diagnosis, "_0318.txt"))[, 1]
}

```

## Functions

#### Fit probability of diagnosis

We fit a two class classifier (diagnosis versus control) using a logistic regression or XGBoost method. Only SNPs known to be associated with a diagnosis is included in the analysis. The intercept of the logistic regression is adjusted to account for biased sampling. For XGBoost, this is adjusted in later steps when calculating case/control probabilities.

```{r single_model_glm}
fit_single_diagnosis_model_glm <- function(data, diagnosis, pheno,
                                       snp_diagnosis_list, 
                                       incidence_rates) {
  
  # Prepare data for a diagnosis vs. control
  data_dx <- data %>%
    filter(DX %in% c(diagnosis, "Control")) %>%
    mutate(dx_binary = as.numeric(DX == diagnosis))
  
  # Get SNP names 
  snp_cols <- colnames(data_dx)[startsWith(colnames(data_dx), "X")]
  
  # Get known diagnosis SNPs
  snp_diagnosis <- snp_diagnosis_list[[diagnosis]]
  
  # Fit logistic regression with known SNPs and the secondary phenotype
  data_selected <- data_dx %>%
    select(all_of(snp_diagnosis), {{pheno}}, dx_binary)
  
  tryCatch({
    fit <- glm(dx_binary ~ ., data = data_selected, family = "binomial")
    
    # Adjust intercept for case-control design
    pi1 <- incidence_rates[[diagnosis]]
    ncase <- sum(data_dx$dx_binary)
    ncontrol <- sum(1 - data_dx$dx_binary)
    adj <- log(ncase * (1 - pi1) / ncontrol / pi1)
    
    coef_vals <- fit$coefficients
    coef_vals[1] <- coef_vals[1] - adj
    
    return(coef_vals)
  }, error = function(e) {
    return(NULL)
  })
}

```

```{r single_model_xgboost}
fit_single_diagnosis_model_xgboost <- function(data, diagnosis, pheno,
                                               snp_diagnosis_list,
                                               incidence_rates,
                                               xgb_params = list(
                                                 max_depth             = 7,
                                                 eta                   = 0.1,
                                                 nrounds               = 400,
                                                 early_stopping_rounds = 10,
                                                 subsample             = 0.5
                                               )) {
  # Prepare binary outcome data
  data_dx <- data %>%
    filter(DX %in% c(diagnosis, "Control")) %>%
    mutate(dx_binary = as.numeric(DX == diagnosis))

  ncase    <- sum(data_dx$dx_binary)
  ncontrol <- sum(1 - data_dx$dx_binary)
  pi0      <- incidence_rates[[diagnosis]] /
              (incidence_rates[[diagnosis]] + incidence_rates[["Control"]])

  # Use only the diagnosis-specific SNPs (instead of all SNP columns)
  snp_diagnosis <- snp_diagnosis_list[[diagnosis]]

  X_data        <- as.matrix(select(data_dx, all_of(snp_diagnosis), all_of(pheno)))
  y_data        <- data_dx$dx_binary
  feature_names <- colnames(X_data)

  # Train / validation split for early stopping
  n_total   <- nrow(X_data)
  n_test    <- ceiling(n_total * 0.2)
  ind_test  <- sample(seq_len(n_total), n_test, replace = FALSE)
  ind_train <- seq_len(n_total)[-ind_test]

  dtrain <- xgb.DMatrix(data = X_data[ind_train, ], label = y_data[ind_train])
  dtest  <- xgb.DMatrix(data = X_data[ind_test,  ], label = y_data[ind_test])
  evals  <- list(train = dtrain, eval = dtest)

  xgb_model_params <- list(
    max_depth = xgb_params$max_depth,
    eta       = xgb_params$eta,
    subsample = xgb_params$subsample,
    objective = "binary:logistic"
  )

  tryCatch({
    fit_xgboost <- xgb.train(
      params                = xgb_model_params,
      data                  = dtrain,
      nrounds               = xgb_params$nrounds,
      evals                 = evals,
      early_stopping_rounds = xgb_params$early_stopping_rounds,
      verbose               = 0
    )

    return(list(
      model         = fit_xgboost,
      feature_names = feature_names,
      pheno         = pheno,
      ncase         = ncase,
      ncontrol      = ncontrol,
      pi0           = pi0
    ))
  }, error = function(e) {
    cat("    Warning: XGBoost fitting failed for", diagnosis, "\n")
    return(NULL)
  })
}
```

```{r all_diagnosis_glm}
# Fit binary logistic models for every diagnosis
fit_ipw_models_glm <- function(data_train, pheno, snp_diagnosis_list,
                               diagnoses, incidence_rates) {
  diagnosis_models <- list()
  for (diagnosis in diagnoses) {
    if (diagnosis == "Control") next
    diagnosis_models[[diagnosis]] <- fit_single_diagnosis_model_glm(
      data_train, diagnosis, pheno, snp_diagnosis_list, incidence_rates
    )
  }
  return(diagnosis_models)
}
```

```{r all_diagnosis_xgboost}
# Fit one XGBoost model for each diagnosis and given phenotype on training data 
fit_ipw_models_xgboost <- function(data_train, pheno, snp_diagnosis_list,
                                   diagnoses, incidence_rates,
                                   xgb_params = list(
                                     max_depth             = 7,
                                     eta                   = 0.1,
                                     nrounds               = 400,
                                     early_stopping_rounds = 10,
                                     subsample             = 0.5
                                   )) {
  diagnosis_models <- list()
  for (diagnosis in diagnoses) {
    if (diagnosis == "Control") next
    diagnosis_models[[diagnosis]] <- fit_single_diagnosis_model_xgboost(
      data_train, diagnosis, pheno, snp_diagnosis_list, incidence_rates, xgb_params
    )
  }
  return(diagnosis_models)
}
```

#### Compute case/control probability

```{r pselect_glm}
# Compute P(D = d | X, Y = y) for every row of X and every diagnosis using GLM
#
# X           : n x p numeric matrix with named columns
# Y           : scalar response value (0 or 1)
# params_list : named list of raw named coefficient vectors, one per non-control
#               diagnosis — output of fit_ipw_models_glm()
# diagnoses   : full character vector including "Control"
# pheno       : phenotype column name, used to extract gamma_y from the vector
#
# Returns     : n x length(diagnoses) matrix of normalised probabilities
pselect_multi <- function(X, Y, params_list, diagnoses, pheno) {
  non_control_dx <- diagnoses[diagnoses != "Control"]
  n              <- nrow(X)

  exp_etas <- matrix(0, nrow = n, ncol = length(non_control_dx))
  colnames(exp_etas) <- non_control_dx

  for (diagnosis in non_control_dx) {
    p_dx <- params_list[[diagnosis]]

    gamma_0 <- p_dx["(Intercept)"]
    gamma_y <- p_dx[pheno]
    # Everything except the intercept and phenotype is a SNP coefficient
    gamma_x <- p_dx[setdiff(names(p_dx), c("(Intercept)", pheno))]

    # Align to X columns; fill any missing coefficients with 0
    gamma_x_matched <- gamma_x[colnames(X)]
    gamma_x_matched[is.na(gamma_x_matched)] <- 0

    v <- Y * gamma_y + X %*% gamma_x_matched + gamma_0

    exp_etas[, diagnosis] <- exp(as.numeric(v))
  }

  denom     <- 1 + rowSums(exp_etas)
  all_probs <- cbind(1 / denom, exp_etas / denom)
  colnames(all_probs)[1] <- "Control"
  all_probs[, diagnoses]
}
```

```{r calibrate_probs_xgboost}
# Calibrate case-control probabilities from XGBoost output to account for biased sampling
#
# prob_xgb : length-n vector of raw XGBoost output probabilities
# ncase    : number of cases in the training set       
# ncontrol : number of controls in the training set    
# pi0      : P(D = d) / (P(D = d) + P("control")) in the population
calibrate_xgb_prob <- function(prob_xgb, ncase, ncontrol, pi0) {
  prob_adj <- 1 / (1 + (1 - pi0) / pi0 *
                       (1 - prob_xgb) / prob_xgb *
                       ncase / ncontrol)          
  prob_adj / (1 - prob_adj)                       
}
```

```{r pselect_xgboost}
# Compute P(D = d | X, Y = y) for every row of X and every diagnosis,
# using pre-fitted XGBoost model bundles and the same multinomial normalisation
# as pselect_multi() 
#
# X                : n x p_snp numeric matrix; column names must include all
#                    SNP names used during training (superset is fine)
# Y                : scalar response value (0 or 1)
# diagnosis_models : named list of model bundles from fit_ipw_models_xgboost()
# diagnoses        : full character vector of diagnosis labels (including "Control")
#
# Returns          : n x length(diagnoses) matrix of normalised probabilities
pselect_multi_xgboost <- function(X, Y, diagnosis_models, diagnoses) {
  non_control_dx <- diagnoses[diagnoses != "Control"]
  n              <- nrow(X)

  exp_odds <- matrix(0, nrow = n, ncol = length(non_control_dx))
  colnames(exp_odds) <- non_control_dx

  for (diagnosis in non_control_dx) {
    fit_obj <- diagnosis_models[[diagnosis]]
    if (is.null(fit_obj)) next

    # Reconstruct the exact feature matrix used during training:
    # SNP columns (in training order) + phenotype column filled with Y
    snp_feat <- setdiff(fit_obj$feature_names, fit_obj$pheno)
    X_snps   <- X[, snp_feat, drop = FALSE]
    X_pheno  <- matrix(Y, nrow = n, ncol = 1,
                       dimnames = list(NULL, fit_obj$pheno))
    X_pred   <- cbind(X_snps, X_pheno)[, fit_obj$feature_names, drop = FALSE]

    prob_xgb <- predict(fit_obj$model, xgb.DMatrix(data = X_pred))

    exp_odds[, diagnosis] <- calibrate_xgb_prob(
      prob_xgb, fit_obj$ncase, fit_obj$ncontrol, fit_obj$pi0
    )
  }

  denom     <- 1 + rowSums(exp_odds)
  all_probs <- cbind(1 / denom, exp_odds / denom)
  colnames(all_probs)[1] <- "Control"
  all_probs[, diagnoses]   
}
```

#### Compute IPW weights 

First compute $P(D = d | X, y)/ P(D = "Control" | X, y)$ for each diagnosis and then normalize the probabilities so they sum up to 1.

```{r ipw_glm}
compute_ipw_glm <- function(data_boot, pheno, snp_diagnosis_list, diagnoses,
                                 incidence_rates, all_snp_names) {
  
  
  # Step 1: Fit diagnosis prediction models on this bootstrap sample
  diagnosis_models <- list()
  
  for(diagnosis in diagnoses) {
    if(diagnosis == "Control") next
    
    model_coef <- fit_single_diagnosis_model_glm(data_boot, diagnosis, pheno,
                                             snp_diagnosis_list, 
                                             incidence_rates)
    diagnosis_models[[diagnosis]] <- model_coef
  }
  
  # Step 2: Compute predicted probabilities for each diagnosis
  weights_matrix <- matrix(0, nrow = nrow(data_boot), 
                           ncol = length(diagnoses) - 1)
  colnames(weights_matrix) <- diagnoses[diagnoses != "Control"]
  
  for(diagnosis in diagnoses) {
    if(diagnosis == "Control") next
    
    coef_vals <- diagnosis_models[[diagnosis]]
    
    # design matrix (intercept + snp + secondary phenotype)
    X_matrix <- as.matrix(select(data_boot, all_of(all_snp_names), {{pheno}}))
    X_design <- cbind(1, X_matrix)
    colnames(X_design)[1] <- "(Intercept)"
    
    # Match coefficients to design matrix columns
    coef_matched <- coef_vals[colnames(X_design)]
    coef_matched[is.na(coef_matched)] <- 0
    
    # Compute predicted probability (note: intercept has been adjusted while fitting the model)
    eta <- X_design %*% coef_matched
    prob <- exp(eta)
    
    weights_matrix[, diagnosis] <- as.numeric(prob)
  }
  
  # Step 3: For controls, compute weight as 1 - sum of diagnosis probabilities
  weight_controls <- 1 / (1 + rowSums(weights_matrix))
  
  # Step 4: Combine weights and normalize to sum to 1
  all_weights <- cbind(weight_controls, weights_matrix * weight_controls)
  colnames(all_weights)[1] <- "Control"
  all_weights <- all_weights[, diagnoses]
  
  # Step 5: Adjust P(D = d | X, Y) for incidence rates
  for(diagnosis in diagnoses) {
    all_weights[, diagnosis] <- all_weights[, diagnosis] / 
      incidence_rates[[diagnosis]]
  }
  
  # Step 6: Get final weight based on individual's actual diagnosis
  final_weight <- rep(NA, nrow(data_boot))
  for(i in 1:nrow(data_boot)) {
    diagnosis <- data_boot$DX[i]
    if(diagnosis %in% colnames(all_weights)) {
      final_weight[i] <- all_weights[i, diagnosis]
    } else {
      final_weight[i] <- NA
      cat("Individual",i, "does not have weight for phenotype ", pheno, ".\n")
    }
  }
  
  # if the weight is less than 5% quantile value then use the 5% quantile 
  q_val <- quantile(final_weight, probs = 0.05, na.rm = TRUE)
  final_weight[final_weight<q_val] <- q_val
  
  data_boot$weight <- 1/ final_weight # use inverse probability weighting 
  
  return(data_boot)
}
```

```{r ipw_xgboost}
compute_ipw_xgboost <- function(data_boot, pheno, snp_diagnosis_list,
                                diagnoses, incidence_rates,
                                xgb_params = list(
                                  max_depth             = 7,
                                  eta                   = 0.1,
                                  nrounds               = 400,
                                  early_stopping_rounds = 10,
                                  subsample             = 0.5
                                )) {

  # Step 1: Fit XGBoost diagnosis prediction models on this data
  # Store full bundle (not just model) to access feature_names, ncase, ncontrol, pi0
  diagnosis_models <- list()
  for (diagnosis in diagnoses) {
    if (diagnosis == "Control") next
    diagnosis_models[[diagnosis]] <- fit_single_diagnosis_model_xgboost(
      data_boot, diagnosis, pheno, snp_diagnosis_list, incidence_rates, xgb_params
    )
  }

  # Step 2: Compute predicted probabilities for each diagnosis
  weights_matrix <- matrix(0, nrow = nrow(data_boot),
                           ncol = length(diagnoses) - 1)
  colnames(weights_matrix) <- diagnoses[diagnoses != "Control"]

  for (diagnosis in diagnoses) {
    if (diagnosis == "Control") next

    fit_obj <- diagnosis_models[[diagnosis]]
    if (is.null(fit_obj)) next

    # Use diagnosis-specific SNPs via feature_names stored in the bundle,
    # ensuring column order exactly matches what was used during training
    X_pred   <- as.matrix(select(data_boot, all_of(fit_obj$feature_names)))
    prob_xgb <- predict(fit_obj$model, xgb.DMatrix(data = X_pred))

    # Calibrate using ncase/ncontrol/pi0 stored in the bundle
    weights_matrix[, diagnosis] <- calibrate_xgb_prob(
      prob_xgb, fit_obj$ncase, fit_obj$ncontrol, fit_obj$pi0
    )
  }

  # Step 3: Compute the weight for controls
  weight_controls <- 1 / (1 + rowSums(weights_matrix))

  # Step 4: Combine weights and normalize to sum to 1
  all_weights <- cbind(weight_controls, weights_matrix * weight_controls)
  colnames(all_weights)[1] <- "Control"
  all_weights <- all_weights[, diagnoses]

  # Step 5: Adjust P(D = d | X, Y) for incidence rates
  for (diagnosis in diagnoses) {
    all_weights[, diagnosis] <- all_weights[, diagnosis] /
                                incidence_rates[[diagnosis]]
  }

  # Step 6: Get final weight based on individual's actual diagnosis
  final_weight <- vapply(seq_len(nrow(data_boot)), function(i) {
    dx <- data_boot$DX[i]
    if (dx %in% colnames(all_weights)) all_weights[i, dx] else NA_real_
  }, numeric(1))

  # Winsorise at 5th percentile
  q_val <- quantile(final_weight, probs = 0.05, na.rm = TRUE)
  final_weight[final_weight < q_val] <- q_val

  data_boot$weight <- 1 / final_weight
  return(data_boot)
}

```

#### BH method

```{r}
# compute threshold for BH method
bh_thresh <- function(pval, q){
  p <- length(pval)
  sorted_pval <- sort(pval) 
  i0 <- max(sum(sorted_pval < q * (1:p) / p), 1) 
  tau <- q * i0 / p # BH-q threshold
  
  return(tau)
}
```

### 

#### Compute conditional distribution of the knockoffs

Given the diagonal matrix $D$, the mean $\mu_\textrm{pop}$ and covariance $\Sigma_\textrm{pop}$, we now compute the conditional distribution of the knockoff $\tilde{X}$ given observed data $X$ using the function below. The formula is given in Gimenez and Zou (2019):

$$
(\tilde{X}^1,\ldots, \tilde{X}^m) \,|\, X \sim N((\tilde{\mu},\ldots, \tilde{\mu}), \tilde{\Sigma}),
$$ where

$$
\tilde{\Sigma} = \left(\begin{matrix}
C & C - D  & C - D \\
C - D & C & C - D\\
C - D & C - D & C
\end{matrix}\right),
\quad 
C = 2D - D\Sigma^{-1}D. 
$$

```{r fun_get_knockoff_mean_cov}
# INPUT
# X -- the data matrix 
# cov -- the covariance matrix
# mu -- the vector of the means
# diag -- the solution of the optimization problem 
# 
# OUTPUT
# mu_tilde -- mean of the knockoffs 
# r_tilde -- covariance matrix for the knockoffs 
knockoff_mean_cov <- function(X, cov, mu, sol){
  m <- 3 # 3 multiple knockoffs
  sd <- sqrt(diag(cov)) # compute std
  SigmaInv <- solve(cov)
  p <- ncol(cov)
      
  sol <- sol * sd^2 # convert the solution back to the covariance 
      
  mu_tilde <- list()
  r_tilde <- list()
  for(i in 1:2){ # use the ME and MVR knockoffs 
    # compute the adjusted matrix Sigma-tilde 
    D <- diag(sol[,i])
    C <- 2 * D - D %*% (SigmaInv %*% D)
    
    # 3 multiple knockoffs 
    SigmaTilde <- rbind(cbind(C, C - D, C - D), cbind(C - D, C , C-D), cbind(C - D, C - D, C))
    # Compute the eigenvalues, if there's an eigen value that is less than eps, print that value out, and turn it to eps
    svd <- eigen(SigmaTilde)
    eigenvalue <- svd$values
    eps <- 0.0005
    if(min(eigenvalue) < eps){
      cat(eigenvalue[which(eigenvalue < eps)])
      eigenvalue[which(eigenvalue < eps)] <- eps
      SigmaTilde <- svd$vectors %*% (diag(eigenvalue) %*% t(svd$vectors))
    }
    
    # store the adjusted covariance matrix 
    RTilde <- chol(SigmaTilde)
        
    # compute the mean term 
    MuTilde0 <- as.vector(D %*% (SigmaInv %*% mu)) + (diag(p) - D %*% SigmaInv) %*% t(X)
    MuTilde <- do.call(rbind, replicate(m, MuTilde0, simplify=FALSE))
    
    mu_tilde[[i]] <- MuTilde
    r_tilde[[i]] <- RTilde
  }
  
  return(list(mu_tilde = mu_tilde, r_tilde = r_tilde))
}
```

#### Functions for knockoffs selection

We fit a regularized logistic regression and use the largest $\lambda$ such that a variable enters the model as the test statistics. Another option is the absolute value of the coefficient at the most regularized model such that the cross-validated error is within one standard error of the minimum. We select variables using the algorithm described in Gimenez and Zou (2019).

```{r fun_lasso_lambda_max_bin}
# Compute lasso_lambda_max statistics 
# INPUT:
# X -- the covariate matrix of size n * p
# Xtilde -- the knockoffs of matrix size n * (3 * p)
# Y -- response variable in 0 or 1
# nlambda -- number of regularization parameters
# OUTPUT:
# the largest lambda where each variable shows up (to get nonzero coef)
lasso_lambda_max_bin <- function(X, XTilde, Y, nlambda = 2e3){
  # fit lasso
  fit <- glmnet(x = cbind(X, XTilde), y = Y, family = "binomial", nlambda = nlambda, intercept = T, standardize = T)
  
  # extract where each coefficient is not zero
  first_nonzero <- function(x) match(T, abs(x) > 0) # NA if all(x==0)
  indices <- apply(fit$beta, 1, first_nonzero)
  # return the list of lambda_max  
  ifelse(is.na(indices), 0, fit$lambda[indices])
}
```

```{r fun_lasso_coef_bin}
# Compute absolute value of the lasso coef 
# INPUT:
# X -- the covariate matrix of size n * p
# Xtilde -- the knockoffs of matrix size n * (3 * p)
# Y -- response variable in 0 or 1
# OUTPUT:
# vector of lasso coef at the lambda that is 1 std away from the value that minimizes the CV error
lasso_coef_bin <- function(X, XTilde, Y){
  # fit lasso
  cvfit <- cv.glmnet(x = cbind(X, XTilde), y = Y, family = "binomial", 
                     intercept = T, standardize = T, type.measure = "class")
  
  # return coefficients (remove the intercept)
  coef(cvfit, s = "lambda.1se")[-1]
}
```

```{r fun_multiple_knockoff_selection}
# Select variables using multiple knockoffs 
# INPUT:
# X -- the covariate matrix
# XTilde -- the knockoff matrix
# Y -- response
# stat -- statistics to use for Tj, "lasso_lambda_max_bin" or "lasso_coef"
# OUTPUT:
# a list of selections at a vector of FDR levels q 
select_vars_multi_knockoff <- function(X, XTilde, Y, kappa = 3, q, stat = "lasso_lambda_max_bin"){
  if(stat == "lasso_lambda_max_bin"){
    stat <- lasso_lambda_max_bin(X, XTilde, Y, nlambda = 2e3)
  }else if(stat == "lasso_coef"){
    stat <- lasso_coef_bin(X, XTilde, Y)
  }
  
  stat <- matrix(stat, nrow = kappa + 1, byrow = T) # arrange into rows (one row is one copy of the original covariate or knockoff)
  stat_sorted <- apply(stat, 2, function(t) sort(t, decreasing = T))
  tau <- stat_sorted[1, ] - stat_sorted[2, ]
  
  sorted_tau <- sort(tau)
  kappa_val <- apply(stat, 2, which.max) == 1
  
  fdp_hat <- sapply(sorted_tau, 
                    function(t) {
                      (1 / kappa + 1/kappa * sum((tau >= t) * (1-kappa_val))) / max(1, sum((tau >= t)*kappa_val))
                    }
  )

  if(min(fdp_hat) <= q){
    threshold <- sorted_tau[min(which(fdp_hat <= q))]
  }else{
    threshold <- NA
  }

  if(!is.na(threshold)){
    selected <- which(tau >= threshold & kappa_val == 1)
  }else{
    selected <- numeric(0)
  }
  
  stat_diff <- stat[1, ] - apply(stat[-1, ], 2, max)
  return(
    list(selected = selected,
         stat_diff = stat_diff)
  )
}
```

#### Sample from a Gaussian distribution

```{r}
sample_gaussian <- function(n, mu = 0, R){
  p <- nrow(R)
  
  X <- matrix(rnorm(n*p, 0, 1), n, p) %*% R
  
  t(t(X) + mu)
}
```

## Method 1: IPW + BH

```{r ipw_bh}
weight_methods <- c("glm", "xgboost")

# Create a list to store results
results_list <- list()

for(pheno in phenos) {
  for(wm in weight_methods) {
    
    # Create a unique name for this combination
    combo_name <- paste(pheno, wm, sep = "_")
    
    # fit GLM + IPW
    data_all <- read.table(paste0(data_dir, "/data/all_data_", 
                                    pheno, "_08092026.txt"),
                             header = TRUE, stringsAsFactors = FALSE)
    all_snp_names <- colnames(data_all)[startsWith(colnames(data_all), "X")]
    
    if(wm == "glm") {
      data_all <- compute_ipw_glm(data_all, pheno, snp_diagnosis_list, 
                                   diagnoses, incidence_rates, all_snp_names)
    } else if(wm == "xgboost") {
      data_all <- compute_ipw_xgboost(data_all, pheno, snp_diagnosis_list, 
                                       diagnoses, incidence_rates, 
                                       xgb_params = xgb_params)
    }
    
    X <- as.matrix(select(data_all, all_of(all_snp_names)))
    y <- pull(data_all, {{pheno}})
    w <- data_all$weight
    
    # Remove rows with missing values
    valid_rows <- !is.na(y) & !is.na(w) & is.finite(w)
    X <- X[valid_rows, ]
    colnames(X) <- snp_names
    y <- y[valid_rows]
    w <- w[valid_rows]
    
    # Fit weighted logistic regression
    fit <- glm(y ~ X, family = "binomial", weights = w)
    # Use robust estimator for SE 
    beta_hat <- coef(fit)
    robust_vcov <- vcovHC(fit, type = "HC0")
    robust_summary <- coeftest(fit, vcov. = robust_vcov)
    
    # compute selections using BH 
    tau <- bh_thresh(robust_summary[-1,4], q = q)
    selected_ipw <- which(robust_summary[-1, 4] <= tau)
    
    # Store results in list
    results_list[[combo_name]] <- selected_ipw
  }
}
```

## Method 2: Standard knockoffs

In order to construct second-order knockoffs, we need to estimate the population mean $\mu_\mathrm{pop}$ and covariance matrix $\Sigma_\mathrm{pop}$. We use our sample to compute an estimate of mean $\mu_d$ and covariance $\Sigma_d$ for each diagnosis $d$ and compute a weighted average based on incidence of each diagnosis.

#### Estimate mean and covariance for each diagnosis

We assume the covariance is block-diagonal, that is, SNPs on different chromosomes are independent of each other.

```{r cov_diagnosis}
snp_names <- read.table(paste0(data_dir, "Processed/processed_data/snp_to_keep_0318.txt"), header = F)[,1]

geno_pruned <- read_table(paste0(data_dir, "Processed/processed_data/geno_pruned_0318.txt"))
  

for(diagnosis in diagnoses){
  # which individual has this diagnosis
  dat <- left_join(geno_pruned, dx, by = "PaisaID") %>% 
      filter(DX == {{diagnosis}} ) %>% 
      select(-c(PaisaID, DX))
  
  cov_mat_diag <- list()
  for(chr in 1:22){
    mat <- select(dat, starts_with(paste0("X", chr,"_")))
    
    cov_mat <- cov(mat, mat, use = "everything")
    
    cov_mat_diag[[chr]] <- cov_mat
  }
  
  mu <- data.frame(snp_names = colnames(dat), mu = colMeans(dat))

  cov <- as.matrix(bdiag(cov_mat_diag))
  colnames(cov) <- snp_names
  rownames(cov) <- snp_names
  
  
  write.table(mu, paste0(data_dir, "Processed/vcov/mu_", diagnosis, "_0318.txt"), row.names = T, col.names = T)
  write.table(cov, paste0(data_dir, "Processed/vcov/cov_", diagnosis, "_0318.txt"), row.names = T, col.names = T)
}
```

#### Population mean and covariance

We assume here that the incidence of BP1 and BP2 as equal, noting that this estimate is not correct.

```{r pop_mean_cov}
for(diagnosis in diagnoses){
  assign(paste0("mu_", diagnosis), read.table(paste0(data_dir, "Processed/vcov/mu_", diagnosis, "_0318.txt"), header = T)$mu)
  
  assign(paste0("cov_", diagnosis), read.table(paste0(data_dir, "Processed/vcov/cov_", diagnosis, "_0318.txt"), header = T))
}

mu_pop <- mu_BP1 * 219 / 100000 / 2 + mu_BP2 * 219 / 100000 / 2 + mu_MDD * 235 / 100000 + mu_SCZ * 45 / 100000 + mu_Control * (100000 - 219 - 235 - 45) / 100000

cov_pop <- cov_BP1 * 219 / 100000 / 2 + cov_BP2 *  219 / 100000 / 2 + cov_MDD * 235 / 100000 + cov_SCZ *  45 / 100000 + cov_Control *  (100000 - 219 - 235 - 45) / 100000
```

```{r store_pop_mean_cov}
write.table(mu_pop, paste0(data_dir, "Processed/vcov/mu_pop_0318.txt"), row.names = F, col.names = T)
  
write.table(cov_pop, paste0(data_dir, "Processed/vcov/cov_pop_0318.txt"), row.names = T, col.names = T)

sd_pop <- sqrt(diag(cov_pop))
corr_pop <- cov_pop / (sd %*% t(sd))

write.table(corr_pop, paste0(data_dir, "Processed/vcov/corr_pop_0318.txt"), row.names = F, col.names = F)
```

```{r compute_pop_corr}
for(diagnosis in diagnoses){
  cov <- read.table(paste0(data_dir, "Processed/vcov/cov_", diagnosis, "_0318.txt")) %>% as.matrix()
  sd <- sqrt(diag(cov))
  corr <- cov / (sd %*% t(sd))
  write.table(corr, paste0(data_dir, "Processed/vcov/corr_", diagnosis, "_0318.txt"), row.names = F, col.names = F)
}
```

#### Solve for the diagonal matrix for each covariance matrices

The covariance matrix of the multiple knockoff has form

$$
\Sigma_{m} = \left(\begin{matrix}
\Sigma & \Sigma - D & \ldots & \Sigma - D \\
\Sigma - D & \Sigma & \ldots & \Sigma - D \\
\vdots & \vdots & \ddots & \vdots \\
\Sigma-D & \Sigma - D & \ldots & \Sigma
\end{matrix}\right)
$$ where $D$ is a diagonal matrix which ensures low correlation between a variable and its knockoff. We use two approaches to compute $D$, the MVR knockoff (see article [Powerful knockoffs via minimizing reconstructability](https://lucasjanson.fas.harvard.edu/papers/Powerful_Knockoffs_Via_Minimizing_Reconstructability-Spector_Janson-2020.pdf) by Spector and Janson) and the ME knockoff (see article [Improving the stability of the knockoff procedure: Multiple Simultaneous knockoffs and entropy maximization](https://proceedings.mlr.press/v89/gimenez19b/gimenez19b.pdf) by Gimenez and Zou). We first convert the covariance to correlation matrix and solve the optimization problem as described in the two articles.

```{julia solve_pop_diag}
# load julia packages
using DelimitedFiles
using CSV
using DataFrames
using Knockoffs

dir = "/scratch/users/qzhao1/paisa/real/data/vcov/"
out_dir = "/scratch/users/qzhao1/paisa/real/data/sol/" 

m = 3; # multiple knockoffs

# load the estimated mu and covariance matrices 
Σ = readdlm(dir * "corr_pop_0318.txt",' ', Float64, '\n');
p = size(Σ )[1]
  
# create some random X
X = randn(100, p);
  
μ = zeros(p);
  
# maximum entropy knockoff
me_multiple = modelX_gaussian_knockoffs(X, :maxent, μ , Σ, m=m);
# mvr knockoff
mvr_multiple = modelX_gaussian_knockoffs(X, :mvr, μ, Σ, m = m);
# equal correlation knockoff 
equi_multiple = modelX_gaussian_knockoffs(X, :equi, μ, Σ, m = m);
  
# concatenate solutions into a matrix
sol = hcat(me_multiple.s, mvr_multiple.s, equi_multiple.s);
  
# save the diagonal terms 
writedlm(out_dir * "sol_pop_0318.txt", sol)
```

#### Sample knockoffs from the Gaussian distribution

Next, we sample $B$ knockoffs (each of these is a multiple knockoff with 3 knockoffs) from the conditional distribution specified above.

```{r}
for(pheno in phenos){
  for(method in c("me", "mvr")){
    cat(pheno, method,  "\n")
    
    r_tilde <- read.table(paste0(data_dir, "ProcessedData/TiltedCov/r_tilde_", pheno, "_",method,"_0508.txt")) %>% as.matrix()
    
    mu_tilde <-  as.matrix(read.table(paste0(data_dir, "ProcessedData/TiltedCov/mu_tilde_", pheno, "_",method,"_0508.txt")))
    
    # sample multivariate Gaussian
    p <- nrow(r_tilde)
    n <- ncol(mu_tilde)
    
    for(b in 1:B){
      cat(b, ",")
      if(b %% 10 == 0) cat("\n")
      xtilde <- matrix(rnorm(n * p), n, p) %*% r_tilde + t(mu_tilde)
      write.table(xtilde, paste0(data_dir, "ProcessedData/Knockoffs/xtilde_", pheno, "_",method,"_", b, ".txt"))
    }
  }
}
```

#### Select variables

```{r selection_standard}
selection_unadjusted <- list()
stat_unadjusted <- list()

snp_names <- read.table(paste0(data_dir, "ProcessedData/Data/snp_to_keep_0318.txt"), header = F)[,1]
for(pheno in phenos){
  data <- read.table(paste0(data_dir, "ProcessedData/PhenoData/", pheno, "_0318.txt"), header = T) %>% filter(.data[[pheno]] != -9)  
  
  X <- select(data, -c(PaisaID, DX, {{pheno}}))
  Y <- pull(data, {{pheno}})
  
  
  for(method in c("me", "mvr")){
    cat(pheno, method, ":\n")
    
    new_selections <- list()
    new_stats <- list()
    for(b in 1:B){
      XTilde <- read.table(paste0(data_dir, "ProcessedData/Knockoffs/xtilde_", pheno, "_",method,"_", b, ".txt")) %>% as.matrix()
      
      new_selection <- select_vars_multi_knockoff(X, XTilde, Y, kappa = 3, q, stat = "lasso_lambda_max_bin")
      new_selections[[b]] <- snp_names[new_selection$selected]
      new_stats[[b]] <- new_selection$stat 
      
      cat(b, ":", snp_names[new_selection$selected], "\n")
    }
    
    new_name <- paste0(pheno, "_", method)
    selection_unadjusted[[new_name]] <- new_selections
    stat_unadjusted[[new_name]] <- new_stats
  }
}

# save results 
saveRDS(selection_unadjusted, file = paste0(data_dir, "ProcessedData/Selection/knockoff.rds"))
saveRDS(stat_unadjusted, file = paste0(data_dir, "ProcessedData/Selection/stat.rds"))
```

## Method 3: Case/control knockoffs

#### Solve for the diagonal matrix for each covariance matrices

Since we generate knockoffs for each diagnoses, we need to solve for the diagonal elements separately for each diagnosis.

```{r corr_pop_diagnosis}
cov_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/ProcessedData/vcov/"
sol_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/ProcessedData/sol/"
cov_tilted_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/ProcessedData/TiltedCov/"
for(diagnosis in diagnoses){
  cov <- read.table(paste0(cov_dir, "cov_", diagnosis, ".txt")) %>% as.matrix()
  sd <- sqrt(diag(cov))
  corr <- cov / (sd %*% t(sd))
  
  write.table(corr, paste0(cov_dir, "corr_", diagnosis,".txt"), row.names = F, col.names = F)

}
```

```{julia}
# load julia packages
using DelimitedFiles
using CSV
using DataFrames
using Knockoffs

diagnoses = ["BP1", "BP2", "SCZ","MDD","Control"]
phenos = ["EID_A_Binary", "EID_S_Binary", "ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary"]

dir = "/work/pi_qianzhao_umass_edu/TiltedKnockoff/ProcessedData/vcov/"
out_dir = "/work/pi_qianzhao_umass_edu/TiltedKnockoff/ProcessedData/sol/"

m = 3; # multiple knockoffs

for diagnosis in diagnoses
print(diagnosis)
# load the estimated mu and covariance matrices 
Σ = readdlm(dir * "corr_" * diagnosis * ".txt",' ', Float64, '\n');
p = size(Σ )[1]

# create some random X
X = randn(100, p);

μ = zeros(p);

# maximum entropy knockoff
me_multiple = modelX_gaussian_knockoffs(X, :maxent, μ , Σ, m=m);
# mvr knockoff
mvr_multiple = modelX_gaussian_knockoffs(X, :mvr, μ, Σ, m = m);
# equal correlation knockoff 
equi_multiple = modelX_gaussian_knockoffs(X, :equi, μ, Σ, m = m);

# concatenate solutions into a matrix
sol = hcat(me_multiple.s, mvr_multiple.s, equi_multiple.s);

# save the diagonal terms 
writedlm(out_dir* "sol_" * diagnosis * ".txt", sol)
end
```

#### Compute conditional distribution of the knockoffs

```{r case_control_knockoff_conditional}
for(diagnosis in diagnoses){
  for(pheno in phenos){
      cat(diagnosis, "_",pheno,"\n")
      
      observed_data <- read.table(paste0("/home/qianzhao_umass_edu/Data/TiltedKnockoff/data/all_data_", 
                                     pheno, "_08092026.txt"),
                              header = TRUE, stringsAsFactors = FALSE)
      X <- observed_data %>% filter(DX == diagnosis) %>% 
       select(-c({{pheno}}, DX, PaisaID)) %>% 
       as.matrix()
     
      # load the covariance matrix
      cov <- read.table(paste0(cov_dir, "cov_", diagnosis, ".txt")) %>% as.matrix()
      
      # load the means
      mu <- read.table(paste0(cov_dir, "mu_", diagnosis, ".txt"), header = T)$x
  
      # load the solved diagonal matrix 
      sol <- read.table(paste0(sol_dir, "sol_", diagnosis,".txt"))
     
      mu_cov_tilded <- knockoff_mean_cov(X, cov, mu, sol)
      
    write.table(mu_cov_tilded$mu_tilde[[1]], paste0(cov_tilted_dir, "mu_tilde_", pheno, "_", diagnosis, "_me_08172026.txt" ) )
    write.table(mu_cov_tilded$mu_tilde[[2]], paste0(cov_tilted_dir, "mu_tilde_", pheno, "_", diagnosis, "_mvr_08172026.txt" ) )
    
    write.table(mu_cov_tilded$r_tilde[[1]], paste0(cov_tilted_dir, "r_tilde_", pheno, "_", diagnosis,"_", "_me_08172026.txt" ) )
    write.table(mu_cov_tilded$r_tilde[[2]], paste0(cov_tilted_dir, "r_tilde_", pheno, "_", diagnosis,"_",  "_mvr_08172026.txt" ) )
  }
}

```

#### Sample knockoffs from the Gaussian distribution

```{r sample_case_control_KO}
# Case/control knockoffs 
for(diagnosis in diagnoses){
  for(pheno in phenos){
    for(sol_method in c("me", "mvr")){
      cat(diagnosis, pheno, sol_method,  "\n")
      
      r_tilde <- read.table(paste0(data_dir, "ProcessedData/TiltedCov/r_tilde_",pheno , "_", diagnosis, "__",sol_method,"_08172026.txt")) %>% as.matrix()
      
      mu_tilde <-  as.matrix(read.table(paste0(data_dir, "ProcessedData/TiltedCov/mu_tilde_", pheno , "_", diagnosis, "_",sol_method,"_08172026.txt")))
      
      # sample multivariate Gaussian
      p <- nrow(r_tilde)
      n <- ncol(mu_tilde)
      
      for(b in 1:B){
        cat(b, ",")
        if(b %% 10 == 0) cat("\n")
        xtilde <- matrix(rnorm(n * p), n, p) %*% r_tilde + t(mu_tilde)
        write.table(xtilde, paste0(data_dir, "ProcessedData/Knockoffs/xtilde_", pheno , "_", diagnosis, "_",sol_method, "_", b, ".txt"))
      }
    }
  }
}
```

#### Select variables

```{r selection_case_contorl_ko_cluster}
# R script run on the cluster
# Case/control knockoff selections 
rm(list = ls())
args <- commandArgs(trailingOnly = TRUE)
ind <- as.numeric(args[1])

cat("ind = ", ind , "\n")

library(rlang, lib.loc = "/home/qianzhao_umass_edu/R/x86_64-pc-linux-gnu-library/4.5/")
library(tidyverse)
library(glmnet)
source("/home/qianzhao_umass_edu/Research/TiltedKnockoff/Real/src.R")

# directory for putting the data 
data_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/"

diagnoses <- c("BP1", "BP2", "SCZ","MDD", "Control")
phenos <- c("EID_A_Binary", "EID_S_Binary", "ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary")

# set number of multiple knockoffs
B <- 50
max_attempts <- B


# FDR threshold 
q <- 0.3 

# Tilted Knockoff 

selection_case_control <- list()
stat_case_control <- list()

snp_names <- read.table(paste0(data_dir, "ProcessedData/Data/snp_to_keep_0318.txt"), header = F)[,1]

for(pheno in phenos[ind]){
  cat(pheno, "\n")
  
  data <- read.table(paste0("/home/qianzhao_umass_edu/Data/TiltedKnockoff/data/all_data_", 
                            pheno, "_08092026.txt"),
                     header = TRUE, stringsAsFactors = FALSE)
  
  for(sol_method in c("me", "mvr")){
    cat(pheno, sol_method, ":\n")
    
    new_selections <- list()
    new_stats <- list()
    
    X <- NULL
    Y <- NULL
    for(diagnosis in diagnoses){
      dat_new <- data %>% 
        filter(DX == diagnosis) %>% 
        select(-c(PaisaID, DX))
      
      y_new <- pull(dat_new, {{pheno}})
      Y <- c(Y, y_new)
      x_new <- select(dat_new, -{{pheno}})
      X <- rbind(X, x_new)
    }
    
    X <- as.matrix(X)
    
    # Modified loop with error handling - continues until 10 successful iterations
    successful_iterations <- 0
    b <- 1
     
    while(successful_iterations < 10 && b <= max_attempts){
      tryCatch({
        XTilde <- NULL
        for(diagnosis in diagnoses){
          xtilde_file <- paste0(data_dir, "ProcessedData/Knockoffs/xtilde_", pheno, "_", 
                                diagnosis, "_", sol_method, "_", b, ".txt")
          xtilde_new <- read.table(xtilde_file)
          XTilde <- rbind(XTilde, xtilde_new)
        }
        XTilde <- as.matrix(XTilde)
        new_selection <- select_vars_multi_knockoff(X, XTilde, Y, kappa = 3, q, stat = "lasso_lambda_max_bin")
        
        # Store results indexed by successful iteration count (not b value)
        new_selections[[successful_iterations + 1]] <- snp_names[new_selection$selected]
        new_stats[[successful_iterations + 1]] <- new_selection$stat 
        
        cat("Success - b:", b, "- Iteration", successful_iterations + 1, ":", 
            paste(snp_names[new_selection$selected], collapse = ", "), "\n")
        
        successful_iterations <- successful_iterations + 1
        
      }, error = function(e) {
        cat("Error at b =", b, "- Skipping. Message:", conditionMessage(e), "\n")
      })
      
      b <- b + 1
    }
    
    # Check if we reached 10 iterations
    if(successful_iterations < 10){
      cat("WARNING: Only achieved", successful_iterations, "successful iterations out of 10 for",
          pheno, "-", sol_method, "\n")
    }
    
    new_name <- paste0(pheno, "_", sol_method)
    selection_case_control[[new_name]] <- new_selections
    stat_case_control[[new_name]] <- new_stats
  }
}

# save results 
saveRDS(selection_case_control, file = paste0(data_dir, "ProcessedData/Selection/case_control_knockoff_", ind, ".rds"))
saveRDS(stat_case_control, file = paste0(data_dir, "ProcessedData/Selection/case_control_stat_", ind, ".rds"))

```

## Method 4: Tilted knockoffs

#### Estimate the mean and covariance matrix of the tilted distribution

```{r}
# glm
diagnosis_models_glm <- list()
for (pheno in phenos) {
  diagnosis_models_glm[[pheno]] <- fit_ipw_models_glm(
    observed_data, pheno, snp_diagnosis_list, diagnoses, incidence_rates
  )
}

# XGBoost
diagnosis_models_xgb <- list()
for (pheno in phenos) {
  diagnosis_models_xgb[[pheno]] <- fit_ipw_models_xgboost(
    observed_data, pheno, snp_diagnosis_list, diagnoses, incidence_rates
  )
}

# Choose which method to use 
method <- "xgboost"   # or "glm"

for (pheno in phenos) {
  for (y in c(0, 1)) {
    for (diagnosis in diagnoses) {
      cat(pheno, y, diagnosis, "\n")

      mu_hat     <- matrix(NA, B, p)
      mu2_hat    <- array(NA, dim = c(B, p, p))
      probs_sums <- numeric(B)

      for (b in 1:B) {
        cat(b, ",")

        # Draw samples from reference distribution P(X)
        x_new <- sample_gaussian(n = 5000, mu = mu, R = R_pop)
        colnames(x_new) <- names(mu)

        # Compute jointly normalised P(D = d | X, Y = y)
        if (method == "glm") {
          all_probs <- pselect_multi(x_new, y, diagnosis_models_glm[[pheno]], diagnoses, pheno = pheno)
        } else {
          all_probs <- pselect_multi_xgboost(x_new, y, diagnosis_models_xgb[[pheno]], diagnoses)
        }

        # Extract importance weights for the target diagnosis
        probs <- all_probs[, diagnosis]

        probs_sums[b] <- sum(probs)
        mu_hat[b, ]   <- t(probs) %*% x_new / sum(probs)
        mu2_hat[b, ,] <- t(x_new * probs) %*% x_new / sum(probs)
      }

      mu_hat_agg    <- t(probs_sums) %*% mu_hat / sum(probs_sums)
      mu2_hat_agg   <- apply(mu2_hat * probs_sums, c(2, 3), sum) / sum(probs_sums)
      sigma_hat_agg <- mu2_hat_agg - t(mu_hat_agg) %*% mu_hat_agg

      write.table(mu_hat_agg,
        paste0(dir, "vcov/mu_hat_", diagnosis, "_", pheno, "_", y, "_05222025.txt"),
        row.names = F, col.names = F)
      write.table(sigma_hat_agg,
        paste0(dir, "vcov/sigma_hat_", diagnosis, "_", pheno, "_", y, "_05222025.txt"),
        row.names = F, col.names = F)
    }
  }
}
```

We then convert the covariance matrix to correlation matrix for solving the diagonal terms for second-order tilted knockoffs

```{r}
cov_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/ProcessedData/vcov/"

for (method in c("glm", "xgboost")){
  for(diagnosis in diagnoses){
    for(pheno in phenos){
      for(y in c(0, 1)){
        cov <- read.table(paste0(cov_dir, "sigma_hat_", method, "_", diagnosis, "_", pheno, "_", y, "_08172026.txt")) %>% as.matrix()
        sd <- sqrt(diag(cov))
        corr <- cov / (sd %*% t(sd))
        write.table(corr, paste0(cov_dir, "corr_", method, "_", diagnosis, "_", pheno, "_", y, "_08172026.txt"), row.names = F, col.names = F)
      }
    }
  }
}

```

#### Solve for the diagonal matrix for each covariance matrices 

Here, we have one covariance matrix for each tilted distribution corresponding to diagnosis d and secondary phenotype y.

```{julia}
# load julia packages
using DelimitedFiles
using CSV
using DataFrames
using Knockoffs

diagnoses = ["BP1", "BP2", "SCZ","MDD","Control"]
phenos = ["EID_A_Binary", "EID_S_Binary", "ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary"]

dir = "/work/pi_qianzhao_umass_edu/TiltedKnockoff/ProcessedData/vcov/"
out_dir = "/work/pi_qianzhao_umass_edu/TiltedKnockoff/ProcessedData/sol/"

m = 3; # multiple knockoffs

for method in ["glm", "xgboost"]
for diagnosis in diagnoses
for pheno in phenos
for y in [0, 1]

println(method, " ",diagnosis, " ", pheno, " ", y)
flush(stdout)
# load the estimated mu and covariance matrices 
Σ = readdlm(dir * "corr_" * method * "_" * diagnosis *"_" * pheno * "_" * string(y) *"_08172026.txt",' ', Float64, '\n');
p = size(Σ )[1]

# create some random X
X = randn(100, p);

μ = zeros(p);

# maximum entropy knockoff
me_multiple = modelX_gaussian_knockoffs(X, :maxent, μ , Σ, m=m);
# mvr knockoff
mvr_multiple = modelX_gaussian_knockoffs(X, :mvr, μ, Σ, m = m);
# equal correlation knockoff 
equi_multiple = modelX_gaussian_knockoffs(X, :equi, μ, Σ, m = m);

# concatenate solutions into a matrix
sol = hcat(me_multiple.s, mvr_multiple.s, equi_multiple.s);

# save the diagonal terms 
writedlm(out_dir* "sol_" * method * "_" * diagnosis *"_" * pheno * "_" * string(y) * "_08172026.txt", sol)
end
end
end
end
```

#### Compute conditional distribution of the knockoffs

```{r conditional_tited}
for(method in c("glm", "xgboost")){
  for(diagnosis in diagnoses){
  for(pheno in phenos){
    for(y in c(0, 1)){
      cat(method, " ",diagnosis, "_",pheno," ", y,"\n")
      
      observed_data <- read.table(paste0("/home/qianzhao_umass_edu/Data/TiltedKnockoff/data/all_data_", 
                                     pheno, "_08092026.txt"),
                              header = TRUE, stringsAsFactors = FALSE)
      
      X <- observed_data %>% filter(DX == diagnosis, .data[[pheno]] == y) %>% 
       select(-c({{pheno}}, DX, PaisaID)) %>% 
       as.matrix()
     
      # load the covariance matrix
      cov <- read.table(paste0(cov_dir, "sigma_hat_", method, "_", diagnosis, "_", pheno, "_", y, "_08172026.txt"), header = F) %>% as.matrix()
      
      # load the means
      mu <- read.table(paste0(cov_dir, "mu_hat_", method, "_", diagnosis, "_", pheno, "_", y, "_08172026.txt")) %>% as.numeric()
  
      # load the solved diagonal matrix 
      sol <- read.table(paste0(sol_dir, "sol_", method, "_", diagnosis, "_", pheno, "_", y, "_08172026.txt"))
     
      mu_cov_tilded <- knockoff_mean_cov(X, cov, mu, sol)
      
    write.table(mu_cov_tilded$mu_tilde[[1]], paste0(cov_tilted_dir, "mu_tilde_", method, "_", pheno, "_", diagnosis, "_", y,"_me_08172026.txt" ) )
    write.table(mu_cov_tilded$mu_tilde[[2]], paste0(cov_tilted_dir, "mu_tilde_", method, "_", pheno, "_", diagnosis, "_", y, "_mvr_08172026.txt" ) )
    
    write.table(mu_cov_tilded$r_tilde[[1]], paste0(cov_tilted_dir, "r_tilde_",method, "_", pheno, "_", diagnosis, "_", y, "_me_08172026.txt" ) )
    write.table(mu_cov_tilded$r_tilde[[2]], paste0(cov_tilted_dir, "r_tilde_", method, "_", pheno, "_", diagnosis, "_", y, "_mvr_08172026.txt" ) )
    }
  }
}
}

```

#### Sample second-order tilted knockoffs

```{r sample_tilted_knockoffs}
# sample tilted knockoffs 
library(tidyverse)

# directory for putting the data 
data_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/"

diagnoses <- c("BP1", "BP2", "SCZ","MDD", "Control")
phenos <- c("EID_A_Binary", "EID_S_Binary", "ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary")

# set number of multiple knockoffs
B <- 50 

# Tilted knockoffs
for(diagnosis in diagnoses){
  for(pheno in phenos){
    for(y in c(0,1)){
      for(method in c("me", "mvr")){
        cat(diagnosis, pheno, method, y,  "\n")
        
        r_tilde <- read.table(paste0(data_dir, "ProcessedData/TiltedCov/r_tilde_", pheno, "_", diagnosis,  "_", y, "_" ,method,"_05222025.txt")) %>% as.matrix()
        
        mu_tilde <-  as.matrix(read.table(paste0(data_dir, "ProcessedData/TiltedCov/mu_tilde_", pheno, "_", diagnosis,  "_",y, "_", method,"_05222025.txt")))
        
        # sample multivariate Gaussian
        p <- nrow(r_tilde)
        n <- ncol(mu_tilde)
        
        for(b in 1:B){
          cat(b, ",")
          if(b %% 10 == 0) cat("\n")
          
          xtilde <- matrix(rnorm(n * p), n, p) %*% r_tilde + t(mu_tilde)
          write.table(xtilde, paste0(data_dir, "ProcessedData/Knockoffs/xtilde_", diagnosis, "_", pheno, "_", y, "_", method,"_", b, ".txt"))
        }
      }
    }
  }
}


```

#### Select variables

```{r selections_tilted}
# R script run on cluster
# Knockoff selections 
args <- commandArgs(trailingOnly = TRUE)
ind <- as.numeric(args[1])
method <- args[2]

cat("ind = ", ind , "\n")

library(rlang, lib.loc = "/home/qianzhao_umass_edu/R/x86_64-pc-linux-gnu-library/4.5/")
library(tidyverse)
library(glmnet)
source("/home/qianzhao_umass_edu/Research/TiltedKnockoff/Real/src.R")

# directory for putting the data 
data_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/"

diagnoses <- c("BP1", "BP2", "SCZ","MDD", "Control")
phenos <- c("EID_A_Binary", "EID_S_Binary", "ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary")

# set number of multiple knockoffs
B <- 50
max_attempts <- B


# FDR threshold 
q <- 0.3 

# Tilted Knockoff 

selection_tilted <- list()
stat_tilted <- list()

snp_names <- read.table(paste0(data_dir, "ProcessedData/Data/snp_to_keep_0318.txt"), header = F)[,1]
for(pheno in phenos[ind]){
  cat(pheno, "\n")
  
  data <- read.table(paste0("/home/qianzhao_umass_edu/Data/TiltedKnockoff/data/all_data_", 
                            pheno, "_08092026.txt"),
                     header = TRUE, stringsAsFactors = FALSE)
  
  for(sol_method in c("me", "mvr")){
    cat(pheno, sol_method, ":\n")
    
    new_selections <- list()
    new_stats <- list()
    
    X <- NULL
    Y <- NULL
    for(diagnosis in diagnoses){
      for(y in c(0, 1)){
        dat_new <- data %>% 
          filter(DX == diagnosis) %>% 
          filter(.data[[pheno]] == y) %>% 
          select(-c(PaisaID, DX))
        
        y_new <- pull(dat_new, {{pheno}})
        Y <- c(Y, y_new)
        x_new <- select(dat_new, -{{pheno}})
        X <- rbind(X, x_new)
      }
    }
    X <- as.matrix(X)
    
    # Modified loop with error handling - continues until 10 successful iterations
    successful_iterations <- 0
    b <- 1
      
    while(successful_iterations < 10 && b <= max_attempts){
      tryCatch({
        XTilde <- NULL
        for(diagnosis in diagnoses){
          for(y in c(0, 1)){
            xtilde_file <- paste0(data_dir, "ProcessedData/Knockoffs/xtilde_", method, "_",
                                  diagnosis, "_", pheno, "_", y, "_", sol_method,"_", b, ".txt")
            xtilde_new <- read.table(xtilde_file)
            XTilde <- rbind(XTilde, xtilde_new)
          }
        }
        XTilde <- as.matrix(XTilde)
        new_selection <- select_vars_multi_knockoff(X, XTilde, Y, kappa = 3, q, stat = "lasso_lambda_max_bin")
        
        # Store results indexed by successful iteration count (not b value)
        new_selections[[successful_iterations + 1]] <- snp_names[new_selection$selected]
        new_stats[[successful_iterations + 1]] <- new_selection$stat 
        
        cat("Success - b:", b, "- Iteration", successful_iterations + 1, ":", 
            paste(snp_names[new_selection$selected], collapse = ", "), "\n")
        
        successful_iterations <- successful_iterations + 1
        
      }, error = function(e) {
        cat("Error at b =", b, "- Skipping. Message:", conditionMessage(e), "\n")
      })
      
      b <- b + 1
    }
    
    # Check if we reached 10 iterations
    if(successful_iterations < 10){
      cat("WARNING: Only achieved", successful_iterations, "successful iterations out of 10 for",
          pheno, "-", sol_method, "\n")
    }
    
    new_name <- paste0(pheno, "_", sol_method)
    selection_tilted[[new_name]] <- new_selections
    stat_tilted[[new_name]] <- new_stats
  }
}
# save results 
saveRDS(selection_tilted, file = paste0(data_dir, "ProcessedData/Selection/tilted_knockoff_", method, "_",ind, ".rds"))
saveRDS(stat_tilted, file = paste0(data_dir, "ProcessedData/Selection/tilted_stat_", method, "_",ind, ".rds"))
```

####  

## Results

#### Selections using standard knockoffs, case-control knockoffs, and tilted knock

We count the number of times each genetic variants is selected for each phenotype using each knockoffs method.

```{r count_selections_ko}
counts_by_pheno <- list()

for(ind in 1:6){
  # Load all method results
  ko_filename <- paste0(dir, "Selection/knockoff_", ind, ".rds")
  cc_ko_filename <- paste0(dir, "Selection/case_control_knockoff_", ind, ".rds")
  tilted_ko_glm_filename <- paste0(dir, "Selection/tilted_knockoff_glm_", ind, ".rds")
  tilted_ko_xgboost_filename <- paste0(dir, "Selection/tilted_knockoff_xgboost_", ind, ".rds")
  
  data_ko <- readRDS(ko_filename)
  data_cc_ko <- readRDS(cc_ko_filename)
  data_tilted_ko_glm <- readRDS(tilted_ko_glm_filename)
  data_tilted_ko_xgboost <- readRDS(tilted_ko_xgboost_filename)
  
  # Initialize counts tibble with all methods
  counts_new <- tibble(name = snp_names,
                 knockoff = numeric(nsnp),
                 case_control_knockoff = numeric(nsnp),
                 tilted_ko_glm = numeric(nsnp),
                 tilted_ko_xgboost = numeric(nsnp))

  # Count standard knockoff selections
  for(name in names(data_ko)){
    for(j in 1:10){
      ind_ko <- sapply(data_ko[[name]][[j]], function(t) which(snp_names == t))
      if(length(ind_ko) > 0) counts_new[ind_ko, "knockoff"] <- counts_new[ind_ko, "knockoff"] + 1
    }
  }
  
  # Count case-control knockoff selections
  for(name in names(data_cc_ko)){
    for(j in 1:10){
      ind_cc_ko <- sapply(data_cc_ko[[name]][[j]], function(t) which(snp_names == t))
      if(length(ind_cc_ko) > 0) counts_new[ind_cc_ko, "case_control_knockoff"] <- counts_new[ind_cc_ko, "case_control_knockoff"] + 1
    }
  }
  
  # Count tilted knockoff (glm) selections
  for(name in names(data_tilted_ko_glm)){
    for(j in 1:10){
      ind_tilted_ko_glm <- sapply(data_tilted_ko_glm[[name]][[j]], function(t) which(snp_names == t))
      if(length(ind_tilted_ko_glm) > 0) counts_new[ind_tilted_ko_glm, "tilted_ko_glm"] <- counts_new[ind_tilted_ko_glm, "tilted_ko_glm"] + 1
    }
  }
  
  # Count tilted knockoff (xgboost) selections
  for(name in names(data_tilted_ko_xgboost)){
    for(j in 1:10){
      ind_tilted_ko_xgb <- sapply(data_tilted_ko_xgboost[[name]][[j]], function(t) which(snp_names == t))
      if(length(ind_tilted_ko_xgb) > 0) counts_new[ind_tilted_ko_xgb, "tilted_ko_xgboost"] <- counts_new[ind_tilted_ko_xgb, "tilted_ko_xgboost"] + 1
    }
  }
  
  # Filter to only keep SNPs selected by at least one method
  counts_new <- filter(counts_new, knockoff > 0 | case_control_knockoff > 0 | 
                                   tilted_ko_glm > 0 | tilted_ko_xgboost > 0)
  counts_new <- mutate(counts_new, pheno = phenos[ind])
  
  # Store in list indexed by phenotype
  counts_by_pheno[[phenos[ind]]] <- counts_new
}
```

#### Consistency between selection methods 

```{r count_overlap}
# Create overlap matrices for each phenotype
overlap_matrices <- list()

for(pheno in names(selected_by_pheno)){
  
  # Get selected SNPs for knockoff methods from selected_by_pheno
  knockoff_selected <- selected_by_pheno[[pheno]] %>%
    filter(knockoff >= 5) %>% pull(name)
  
  cc_ko_selected <- selected_by_pheno[[pheno]] %>%
    filter(case_control_knockoff >= 5) %>% pull(name)
  
  tko_glm_selected <- selected_by_pheno[[pheno]] %>%
    filter(tilted_ko_glm >= 5) %>% pull(name)
  
  tko_xgb_selected <- selected_by_pheno[[pheno]] %>%
    filter(tilted_ko_xgboost >= 5) %>% pull(name)
  
  # Get selected SNPs for IPW-GLM from results_list
  ipw_glm_selected <- c()

    combo_name <- paste(pheno, "glm", sep = "_")
    if(combo_name %in% names(results_list)){
      # Convert indices to SNP names
      ipw_glm_selected <- c(ipw_glm_selected, snp_names[results_list[[combo_name]]])
    }

  ipw_glm_selected <- unique(ipw_glm_selected)
  
    ipw_xgboost_selected <- c()

    combo_name <- paste(pheno, "xgboost", sep = "_")
    if(combo_name %in% names(results_list)){
      # Convert indices to SNP names
      ipw_xgboost_selected <- c(ipw_xgboost_selected, snp_names[results_list[[combo_name]]])
  }
  ipw_xgboost_selected <- unique(ipw_xgboost_selected)
  
  # Create a list of selected SNPs for each method
  methods <- c("knockoff", "case_control_knockoff", "tilted_ko_glm", "tilted_ko_xgboost", "ipw_glm", "ipw_xgboost")
  selected_lists <- list(
    knockoff = knockoff_selected,
    case_control_knockoff = cc_ko_selected,
    tilted_ko_glm = tko_glm_selected,
    tilted_ko_xgboost = tko_xgb_selected,
    ipw_glm = ipw_glm_selected,
    ipw_xgboost = ipw_xgboost_selected
  )
  
  # Create overlap matrix
  n_methods <- length(methods)
  overlap_matrix <- matrix(0, nrow = n_methods, ncol = n_methods, 
                           dimnames = list(methods, methods))
  
  # Fill the matrix with counts of overlapping selections
  for(i in 1:n_methods){
    for(j in 1:n_methods){
      overlap_matrix[i, j] <- length(intersect(selected_lists[[i]], selected_lists[[j]]))
    }
  }
  
  overlap_matrices[[pheno]] <- overlap_matrix
}
```

Fig.16 in supplement

```{r plot_overlap}
# Define method names
method_names <- c("No adjustment", "Case/control KO", "Tilted KO (glm)", "Tilted KO (XGBoost)", "IPW (glm)", "IPW (XGBoost)")

for(pheno in phenos){
  overlap_matrix_renamed <- overlap_matrices[[pheno]]
  rownames(overlap_matrix_renamed) <- method_names
  colnames(overlap_matrix_renamed) <- method_names
  overlap_matrices[[pheno]] <- overlap_matrix_renamed
}

phenos <- c( "ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary")

fig_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/result/fig/"
# Create heatmaps for all phenotypes and save to PDFs
for(pheno in phenos){
  png(paste0(fig_dir, pheno,".png"), width = 16, height = 12, units="cm", res = 300)
  
  
  pheno_title <- case_when(
    pheno == "ATT_A_Binary" ~ "Attention (Accuracy)",
    pheno == "ATT_S_Binary" ~ "Attention (Speed)",
    pheno == "PS_A_Binary" ~ "Processing speed (Accuracy)",
    pheno == "PS_S_Binary" ~ "Processing speed (Speed)",
    TRUE ~ pheno  # default to phenotype name for EID phenotypes
  )
  pheatmap(overlap_matrices[[pheno]], 
           display_numbers = TRUE,
           cluster_rows = FALSE, 
           cluster_cols = FALSE,
           main = paste(pheno_title),
           cellwidth = 30,
           cellheight = 30,
           fontsize = 12,
           number_format = "%.0f",
           color = colorRampPalette(c("white", "lightblue", "darkblue"))(50),
           border_color = "black",
           border_width = 1,
           angle_col = 45)
  
  dev.off()
  
}
```

Fig.5 in Paper

```{r plot_overall_overlap}
# Define method names
method_names <- c("No adjustment", "Case/control KO", "Tilted KO (glm)", "Tilted KO (XGBoost)", "IPW (glm)")

# Sum all overlap matrices
overlap_matrix_total <- overlap_matrices[[names(overlap_matrices)[3]]] * 0  # Initialize with zeros

for(pheno in phenos){
  overlap_matrix_total <- overlap_matrix_total + overlap_matrices[[pheno]]
}

# Create heatmap for the summed overlaps
png(paste0(fig_dir,"total_overlap.png"), width = 16, height = 12, units="cm", res = 300)
  

pheatmap(overlap_matrix_total, 
         display_numbers = TRUE,
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         # main = "Total Overlap of Selected SNPs Across All Phenotypes",
         cellwidth = 30,
         cellheight = 30,
         fontsize = 12,
         number_format = "%.0f",
         color = colorRampPalette(c("white", "lightblue", "darkblue"))(50),
         border_color = "black",
         border_width = 1,
         angle_col = 45)

dev.off()

```

#### Compute z-scores

```{r compute_z_sccore}
result <- NULL

thresh <- 5
for(ind in 1:4){
  pheno <- phenos[ind]
  dat <- filter(selected_by_pheno[[pheno]]) %>% 
    filter(knockoff >= thresh | tilted_ko_xgboost >= thresh)
  
  selected_snps <- pull(dat, name)
  
  # get data for the corresponding phenotype
  data_all <- read.table(paste0(data_dir, "/data/all_data_", 
                                    pheno, "_08092026.txt"),
                             header = TRUE, stringsAsFactors = FALSE)
  
  all_snp_names <- colnames(data_all)[startsWith(colnames(data_all), "X")]
  
  #  add weights using xgboost 
  data_all <- compute_ipw_xgboost(data_all, pheno, snp_diagnosis_list, 
                                       diagnoses, incidence_rates, 
                                       xgb_params = xgb_params)
    
  
  X <- as.matrix(select(data_all, selected_snps))
  y <- pull(data_all, {{pheno}})
  w <- data_all$weight
  
  # Remove rows with missing values
  valid_rows <- !is.na(y) & !is.na(w) & is.finite(w)
  X <- X[valid_rows, ]
  colnames(X) <- selected_snps
  y <- y[valid_rows]
  w <- w[valid_rows]
  
  # Fit weighted logistic regression
  fit <- glm(y ~ X, family = "binomial", weights = w)
  # Use robust estimator for SE 
  beta_hat <- coef(fit)
  robust_vcov <- vcovHC(fit, type = "HC0")
  robust_summary <- coeftest(fit, vcov. = robust_vcov)
  

  result_new <- bind_cols(dat, robust_summary[-1,3:4, drop = F])
  
  result <- bind_rows(result, result_new)
}
```

Fig.6 in paper

```{r plot_z_score}
result <- result %>% 
  mutate(z = abs(`z value`),
         ratio = knockoff / tilted_ko_xgboost) 

result <- result %>% 
  mutate(Phenotype = case_when(pheno == "ATT_A_Binary"~"Attention (Continuous performance test)",
                               pheno == "ATT_S_Binary"~"Attention (Continuous performance test)",
                               pheno == "PS_A_Binary"~"Processing Speed (Digit symbol test)",
                                pheno == "PS_S_Binary"~"Processing Speed (Digit symbol test)"
                               ),
         Measure = case_when(pheno == "ATT_A_Binary"~"Accuracy",
                          pheno == "ATT_S_Binary"~"Speed",
                          pheno == "PS_A_Binary"~"Accuracy",
                          pheno == "PS_S_Binary"~"Speed"
                                ))

selection <- filter(result, knockoff >= thresh, tilted_ko_xgboost >= thresh)


gg <- ggplot() + 
  geom_point(aes(x = z, y = ratio, color = Phenotype, shape = Measure), data = result, size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed") + 
  geom_point(aes(x = selection$z, y = selection$ratio), size = 6, color = "#fe3b1f", fill = NA, shape = 21, stroke = 1) +
  xlab("Z-score") + 
  ylab("Ratio (KO / Tilted KO)") + 
  scale_color_manual(values = c( "#1D7AE2", "#F7942E")) + 
  # annotate("text", x = 0.35, y = 0.25, label = "highlight?", size = 8)
  theme_bw() + 
  theme(text = element_text(size = 15), 
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12))



ggsave(gg, filename = paste0(fig_dir, "nselect_real.png" ), height = 5, width = 9, unit = "in")

```

#### Number of previously identified SNPs

```{r}
EffectSizeIndexSNPs <- read_delim(paste0(data_dir, "/9KSmall01112024/EffectSizeIndexSNPs.txt"), delim = "\t", escape_double = FALSE, trim_ws = TRUE)

# number of SNPs for each diagnosis
EffectSizeIndexSNPs %>% 
  group_by(GWAS_file) %>% 
  summarize(n())

# number of SNPs that are associated with more than one diagnosis
EffectSizeIndexSNPs %>% 
  group_by(sigSNPS) %>% 
  filter(n() > 1)
```

#### Estimated probability of diagnosis given phenotype and genotype

Fig.12-15 in supplement

```{r}
fig_dir <- "/work/pi_qianzhao_umass_edu/TiltedKnockoff/result/fig/"

r2_cohen <- tibble(diagnosis = character(),
                   pheno = character(),
                   r2 = numeric())
diagnoses <- c("BP1", "BP2", "MDD", "SCZ","Control")
phenos <- c("EID_A_Binary", "EID_S_Binary", "ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary")


for(pheno in c("ATT_A_Binary", "ATT_S_Binary", "PS_A_Binary", "PS_S_Binary")){
    for(diagnosis in c("BP1", "BP2", "MDD", "SCZ")){
    dx <- case_when(diagnosis == "BP1"~"Bipolar disorder I",
                        diagnosis == "BP2"~"Bipolar disorder II",
                        diagnosis == "MDD"~"Major depressive disorder",
                        diagnosis == "SCZ"~"Schizophrenia", 
                        
                               )
    phenotype <-  case_when(pheno == "ATT_A_Binary"~"Continuous performance test (accuracy)",
                        pheno == "ATT_S_Binary"~"Continuous performance test (speed)",
                        pheno == "PS_A_Binary"~"Digit symbol test (accuracy)",
                        pheno == "PS_S_Binary"~"Digit symbol test (speed)", 
                        
                               )

    data <- read.table(paste0(data_dir, "/data/all_data_", 
                                pheno, "_08092026.txt"),
                         header = TRUE, stringsAsFactors = FALSE)
     
      # Prepare data for a diagnosis vs. control
      data_dx <- data %>%
        filter(DX %in% c(diagnosis, "Control")) %>%
        mutate(dx_binary = as.numeric(DX == diagnosis))
      
      # Get SNP names 
      snp_cols <- colnames(data_dx)[startsWith(colnames(data_dx), "X")]
      
      # Get known diagnosis SNPs
      snp_diagnosis <- snp_diagnosis_list[[diagnosis]]
      
      # Fit logistic regression with known SNPs and the secondary phenotype
      data_selected <- data_dx %>%
        select(all_of(snp_diagnosis), {{pheno}}, dx_binary)
      
      fit <- glm(dx_binary ~ ., data = data_selected, family = "binomial")
   
    r2 <- (fit$null.deviance - fit$deviance) / fit$null.deviance
    
    new_r2 <- data.frame(
      diagnosis = diagnosis,
      pheno = pheno,
      r2 = r2
    )
    
    r2_cohen <- bind_rows(r2_cohen, new_r2)
    
    # create histograms of log-odds with contribution from genotype and phenotype 
    g <- tibble(pred = fit$linear.predictors - fit$coef[1],
           pheno = pull(data_dx, {{pheno}})) %>% 
      mutate(pheno = factor(pheno)) %>% 
     ggplot() + 
      geom_histogram(aes(x = pred,  fill = pheno),
                     color = "grey30",
                     alpha = 0.5, position = "identity", bins = 30 ) + 
      scale_fill_manual(values = c("#E1BE6A", "#D81B60")) + 
      xlab("Linear predictor") + 
      ylab("Frequency") + 
      labs(fill = phenotype)  + 
      ggtitle(dx) + 
      # scale_color_manual(values = c("#E1BE6A", "#1AFF1A")) + 
      theme_bw() + 
      theme(text = element_text(size = 18),
            legend.position="bottom")
    
    assign(paste0("g_", diagnosis), g)
    
    }
  g <- ggarrange(g_BP1, g_BP2, g_MDD, g_SCZ, ncol = 2,nrow = 2, common.legend = T, legend="bottom") 
  
  ggsave(g, filename = paste0(fig_dir, pheno, "_linear_predictor.png" ), height = 10, width = 12, unit = "in")
}

# write.table(r2_cohen, file = "/Users/qianzhao/Google Drive/My Drive/Research/TiltedKnockoff/Tilted knockoff/TiltedKnockoff(05142025)/r2.txt", sep = "\t")

```
